DSCI 445 - Algal Blooms

### 2020-12-14 paper.Rmd
cleanup

With the increasing use of reservoirs for municipalities, the quality of the 
water is very important. Although the water gets treated before it gets to the 
tap, the quality of the water as it reaches the water treatment plant is very 
important regarding the cost to treat the water for municipal use, and algal 
blooms impact the quality of the water dramatically(Brookes). Most municipal water is 
monitored, but not all of the data is publicly available and some isn’t 
monitored in situ - otherwise known as collected by hand. However, Landsat is a
series of satellites that image the Earth surface, in repeated intervals of 14 
days; that is to say Fort Collins gets re-imaged every 14 days. Landsat 5 dates
back to 1984, and is no longer taking images, while landsat 7 (1999) and landsat
8 (2013) still record images today. The landsat imagery allows for a larger 
sample of data to look at the trends of algal blooms in Colorado, in comparison 
to the lack of public in situ data available. Figure 1 shows reservoirs - only 
those greater than 0.01 km² -  that have both in situ and satellite imagery in 
blue, while the reservoirs in red indicate the reservoirs with just the landsat 
imagery available. It is easy to see the lack of reservoir data we have 
available to us, which is why utilizing machine learning to try and predict 
algae blooms from just remotely sensed imagery would be an incredibly useful
method of determining water quality in reservoirs. It has been done previously,
but mainly utilizing only the blue and green wavelengths, despite there being more
potential options(Sachidananda).

![](images/map.jpeg)

![](images/HT_Climatology.jpeg)

![](images/ht_color_over_time.jpeg)

@@

The data came from two sources; satellite data and in situ data from public
sources. The satellite data was downloaded utilizing Google Earth Engine, which
downloaded the data for individual lakes, ignoring images that were over a
threshold of cloud cover. This threshold was because cloud cover would darken
the pixels, skewing the data. The satellite data included all of Landsat 5, 7,
and 8 from the past 35 year and has the different band (wavelength) data for the
center most pixel of each lake. The in situ data includes all public data that
is available for chlorophyll in Colorado. This includes data from sources like
Denver Water, the USGS, and Northern Water. The data that is being utilized is
chlorophyll values, and the data also includes secchi depth (how deep you can
see into the water).

@@

Satellite image data and in situ data were not collected on similar schedules.
This meant that many observations were lost from both datasets on an inner join.
To remedy this problem, a procedure was used referred to as a "merge threshold."
Using a merge threshold, the in situ data was shifted forward and backward $0,
1, \cdots, n$ days. For example, on a merge threshold of two, the in situ data
from any single observation on day $k$ would be used on day $k - 2, k - 1, k, k
+ 1$, and $k + 2$. In other words, the data from any single in situ observation
  would be used for the surrounding four days. The merge threshold procedure was
  done up to ten days of threshold.

@@

The algae bloom data is spatiotemporal; the blooms are recorded in a certain
geographic location at a certain time. This makes the class of linear models
unsuitable for this problem. Linear models make assumptions about the structure
of the errors in the data generating process that are violated by spatiotemporal
data. Spatiotemporal data has errors which are correlated; as observations
closer in space or time will covary.

The models selected were tree-based: Random Forest, Boosted Trees, and XGBoost. Although
the classical formulation of Random Forests and Boosted Trees rely on the
assumption of independent and identically distributed observations, tree-based
models were chosen in hopes that the extra flexibility in these models would
help account for the temporal structure in the data. Unfortunately, there was no
explicit spatial variable that could be included in the data to account for
spatial dependence.

The Random Forest and Bosted Tree implementation from `scikit-learn` were used.
The XGBoost implementation was taken from the Python package `xgboost`.

@@

After a training and test set were created, $5$-fold cross-validation was
performed. Models were compared via $5$-fold cross validation error on the
training dataset. The final model was selected via this error, and then results
were repoted on the test data set.

@@

Finally, we generate chlorophyll estimates for each of the available satellite images,
including satellite images of lakes for which there is no available in situ data.
\ref{fig:predicted-values} Shows the estimated chlorophyll value against the dominant
wavelength of the satellite image. We can see that the highest chlorophyll estimates
are attributed to satellite images which have a green dominant wavelength. However, we
are generating predictions of chlorophyll values exceeding $70 \frac{\mu g}{mL}$, which
is most likely unreasonable. This is likely an artifact of the extreme observations included
in the training data with chlorophyll measurements exceeding $300 \frac{\mu g}{mL}$.

@@

In future exploration, we hope to address the extreme predicted values by removing observations with a
chlorophyll concentration exceeding $15 \frac{\mu g}{mL}$. We expect this change to result in a much
more reasonable range of predicted values. We also think that, based on the importance of the lake area
feature, the inclusion of a feature describing lake elevation will be important

>>>>

With the increasing use of reservoirs for municipalities, the quality of the 
water is very important. Although the water gets treated before it gets to the 
tap, the quality of the water as it reaches the water treatment plant is very 
important regarding the cost to treat the water for municipal use, and algal 
blooms impact the quality of the water dramatically(Brookes). Most municipal water is 
monitored, but not all of the data is publicly available and some isn’t 
monitored in situ - otherwise known as collected by hand.

We will use two distinct datasets: satellite data and in situ data from public
sources. The satellite data were downloaded from Google Earth Engine, which
downloaded the data for individual lakes, ignoring images that were over a
threshold of cloud cover. This threshold was because cloud cover would darken
the pixels, skewing the data. The satellite data included all of Landsat 5, 7,
and 8 from the past 35 year and has the different band (wavelength) data for the
center most pixel of each lake. The in situ data includes all public data that
is available for chlorophyll in Colorado. This includes data from sources like
Denver Water, the USGS, and Northern Water. The data that is being utilized is
chlorophyll values, which act as a proxy for algae blooms.

![Observed and unobserved lakes in Colorado. \label{map}](images/map.jpeg)

Landsat is a series of satellites that image the Earth surface, in repeated 
intervals of 14  days; that is to say an individual location such as Fort Collins gets
re-imaged every 14 days. Landsat 5 dates back to 1984, and is no longer taking images,
while landsat 7 (1999) and landsat 8 (2013) still record images today. The landsat
imagery allows for a larger sample of data to look at the trends of algal blooms in
Colorado, in comparison to the lack of public in situ data available. \ref{fig:map}
shows reservoirs - only those greater than 0.01 km² -  that have both in situ and
satellite imagery in blue, while the reservoirs in red indicate the reservoirs with 
just the landsat imagery available. It is easy to see the lack of reservoir data we have 
available to us, which is why utilizing machine learning to try and predict 
algae blooms from just remotely sensed imagery would be an incredibly useful
method of determining water quality in reservoirs. It has been done previously,
but mainly utilizing only the blue and green wavelengths, despite there being more
potential options(Sachidananda).

![Color (dominant wavelength) distributions for Horsetooth Reservoir in each month. \label{climatology}](images/HT_Climatology.jpeg){width=75%}

![Color (dominant wavelength) of Horsetooth Reservoir over time. \label{color-trend}](images/ht_color_over_time.jpeg){width=75%}

We attempted to detect algae blooms from satellite images by using statistical learning
methods to approximate the relationship between features in satellite images and in situ
chlorophyll measurements from various lakes in Colorado (predominantly municipal water
sources such as Horsetooth Reservoir). We reduced the satellite images to a set of
key features that are likely to be associated with algae blooms; red, green, and blue
wavelength range measurements, dominant wavelength, and various infrared range measurements.
We joined these data to in situ chlorophyll measurements and fit various non-parametric
models before finally producing chlorophyll estimates for all available satellite images.

@@

@@

Satellite image data and in situ data were not collected on similar schedules.
This meant that many observations were lost from both datasets on an inner join.
To remedy this problem, a procedure was used referred to as a "merge threshold."
Using a merge threshold, the satellite data were shifted forward and backward $0,
1, \cdots, n$ days. For example, on a merge threshold of two, the in situ data
from any single observation on day $k$ would be used on day $k - 2, k - 1, k, k
+ 1$, and $k + 2$. In other words, the data from any single in situ observation
would be used for the surrounding four days, but each in situ observation would
only be included once. The merge threshold procedure was done up to ten days of
threshold.

@@

The algae bloom data is spatiotemporal; the blooms are recorded in a certain
geographic location at a certain time. This makes the class of linear models
unsuitable for this problem. Linear models make assumptions about the structure
of the errors in the data generating process that are violated by spatiotemporal
data. Spatiotemporal data have errors which are correlated; as observations
closer in space or time will vary together.

The models selected were tree-based: Random Forest, Boosted Trees, and XGBoost. Although
the classical formulation of Random Forests and Boosted Trees rely on the
assumption of independent and identically distributed observations, tree-based
models were chosen in hopes that the extra flexibility in these models would
help account for the temporal structure in the data.

The Random Forest and Boosted Tree implementation from `scikit-learn` were used.
The XGBoost implementation was taken from the Python package `xgboost`.

@@

After a training and test set were created, $5$-fold cross-validation was
performed. Models were compared via $5$-fold cross validation error on the
training dataset. The final model was selected via this error, and then results
were reported on the test data set.

@@

Finally, we generated chlorophyll estimates for each of the available satellite images,
including satellite images of lakes for which there is no available in situ data.
\ref{fig:predicted-values} Shows the estimated chlorophyll value against the dominant
wavelength of the satellite image. We can see that the highest chlorophyll estimates
are attributed to satellite images which have a green dominant wavelength. However, we
are generating predictions of chlorophyll values exceeding $70 \frac{\mu g}{mL}$, which
is most likely unreasonable. This is likely an artifact of the extreme observations included
in the training data with chlorophyll measurements exceeding $300 \frac{\mu g}{mL}$.

@@

In future exploration, we hope to address the extreme predicted values by removing observations with a
chlorophyll concentration exceeding $15 \frac{\mu g}{mL}$. We expect this change to result in a much
more reasonable range of predicted values. We also think that, based on the importance of the lake area
feature, the inclusion of a feature describing lake elevation will be important. Lastly, although we
included indicators for month to account for variance in time, we wish to repeat the analysis including
spatial features such as latitude and longitude to account for variance in space.
